---
title: "Maps"
output: html_document
date: "2024-12-08"
---

check if I still need several values in color vectors or how to undo that

# load packages
```{r}
library(tidyverse)
library(here)
library(sf)
library(ggspatial)
library(rnaturalearth)
library(ggOceanMaps)
```


# read in data
```{r}
byl_raw <- read.csv(here("Polynoidae - Bylgides mus nub=my num=DNA.csv")) |> janitor::clean_names()
```


# make table for locations
```{r}
byl_locations <- byl_raw |>
  select(species, clade, marthas_number, project_number, bold_process_specimen_id, latitude, longitude) |>
  drop_na() |>
  filter(clade != "", clade != "unsampled", clade != "not_on_tree") |> 
  mutate(clade = factor(clade)) |> 
  mutate(type = if_else(grepl("type", species, ignore.case = TRUE), "type", "nontype")) |> #this introduces the type or nontype column
  mutate(type = factor(type))
```


# world distribution map (Mercator projection)
```{r}
world <- ne_countries(scale = 10, returnclass = "sf")

# plot it (run sf objects first for this to work)
ggplot() +
  geom_sf(data = world) +
  geom_sf(
    data = byl_sf,
    aes(color = clade),
    size = 1.5,
    alpha = .8
  ) +
  labs(color = "Clade", x = "Longitude", y = "Latitude", color = "Clade") +
  theme_minimal()
```






# world distribution map (polar projection)
Once you've run this code, some other functions won't work anymore since packages are conflicting. Both restarting R or running some unload-code help:
detach("package:mapview", unload=TRUE)
detach("package:maptools", unload=TRUE)
detach("package:raster", unload=TRUE)
```{r}
# load packages
packages <- list("dplyr", "sf", "ggplot2","mapview", "maptools", "raster")
sapply(packages, require, character = TRUE)

# plotting coastlines of different extent
data("wrld_simpl", package = "maptools")

w_arctic <- raster::crop(wrld_simpl, extent(-180, 180, 30, 90))
w_boreal <- raster::crop(wrld_simpl, extent(-180, 180, 20, 90))
    #plot(w)

# make them into sf objects, transform the crs
w_arctic_sf <- w_arctic |> 
  st_as_sf(coords = c("long", "lat"), crs = 4326) |> 
  st_transform(st_crs(3995))
w_boreal_sf <- w_boreal |> 
  st_as_sf(coords = c("long", "lat"), crs = 4326) |> 
  st_transform(st_crs(3995))



# make function to plot
distribution <- function(mybasesf, mysf, mycol, mylabels, mylegend)
  {ggplot() + 
  geom_sf(data= mybasesf, fill = "black", colour = "white", alpha = 0.6) +
  geom_sf(data = mysf, aes(col = clade, shape = type), size = 2.5, alpha = 1) +
  scale_shape_manual(values = c(19, 17), #19 is a dot, 17 a triangle
                     labels = c("specimen", "type specimen"),
                     guide = "none") + 
  scale_color_manual(values = mycol,
                     labels = mylabels) +
  scale_x_continuous(breaks = NULL) +
  labs(title = "", x = "", y = "") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank(), 
        text = element_text(size = 12),
        legend.title = element_blank(),
        legend.position = mylegend,
        legend.byrow = TRUE)
}

```


# make sf objects
If the code gives weird errors, check the note in the section "world distribution map (polar projection)"
```{r}
# all species (coverage)
byl_sf <- st_as_sf(byl_locations,
                   coords = c("longitude", "latitude"),
                   crs = 4326)

# sensu stricto species
byl_senstr <- byl_locations |> 
  filter(clade != "macrolepidus", clade != "acutisetis", clade != "acutisetis_close", clade != "fuscus", clade != "belfastensis", clade != "bylgides")

byl_senstr_sf <- st_as_sf(byl_senstr,
                   coords = c("longitude", "latitude"),
                   crs = 4326)

# excluded species
byl_excl <- byl_locations |> 
  filter(clade == "macrolepidus" | clade == "acutisetis" | clade == "acutisetis_close" | clade == "fuscus" | clade == "belfastensis" | clade == "bylgides")

byl_excl_sf <- st_as_sf(byl_excl,
                   coords = c("longitude", "latitude"),
                   crs = 4326)

# acutisetis & acutisetis_close
byl_acutisetis <- byl_locations |> 
 filter(clade == "acutisetis" | clade == "acutisetis_close")

byl_sf_acutisetis <- st_as_sf(byl_acutisetis,
                   coords = c("longitude", "latitude"),
                   crs = 4326)

# annenkovae
byl_annenkovae <- byl_locations |> 
  filter(clade == "annenkovae")


byl_sf_annenkovae <- st_as_sf(byl_annenkovae,
                   coords = c("longitude", "latitude"),
                   crs = 4326)

# elegans
byl_elegans <- byl_locations |> 
  filter(clade == "elegans")


byl_sf_elegans <- st_as_sf(byl_elegans,
                   coords = c("longitude", "latitude"),
                   crs = 4326)

# groenlandicus
byl_groenlandicus <- byl_locations |> 
  filter(clade == "groenlandicus")


byl_sf_groenlandicus <- st_as_sf(byl_groenlandicus,
                   coords = c("longitude", "latitude"),
                   crs = 4326)

# promamme
byl_promamme <- byl_locations |> 
  filter(clade == "promamme" | clade == "promamme_close")


byl_sf_promamme <- st_as_sf(byl_promamme,
                   coords = c("longitude", "latitude"),
                   crs = 4326)

# sarsi
byl_sarsi <- byl_locations |> 
  filter(clade == "sarsi")


byl_sf_sarsi <- st_as_sf(byl_sarsi,
                   coords = c("longitude", "latitude"),
                   crs = 4326)
```


# set colors
```{r}
# set colors for the species and their types
colors_senstr <- c("#228834ff", "#f0e442ff", "#d55e00ff", "#7e2954ff","#cc79a7ff","#65dbedff")
colors_excl <- c("#5a8efcff", "#735deeff", "#000049ff", "grey", "#5da899ff", "#ef6678ff")

colors_sampling_coverage <- rep("red", 12) 
colors_acutisetis <- c("#5a8efcff", "#735deeff")
colors_annenkovae <- c("#228834ff")
colors_belfastensis <- c("#000049ff")
colors_bylgides <- c("grey")
colors_elegans <- c("#f0e442ff")
colors_fuscus <- c("#5da899ff")
colors_groenlandicus <- c("#d55e00ff")
colors_macrolepidus <- c("#ef6678ff")
colors_promamme <- c("#7e2954ff","#cc79a7ff")
colors_sarsi <- c("#65dbedff")

```


# plot species distributions
```{r}
distr_coverage <- distribution(mybasesf = w_boreal_sf,
                               mysf = byl_sf,
                               mycol = colors_sampling_coverage,
                               mylabels = "none",
                               mylegend = "none")
distr_coverage

distr_senstr <- distribution(mybasesf = w_arctic_sf,
                          mysf = byl_senstr_sf,
                          mycol = colors_senstr,
                          mylabels = c(
        expression(italic("B. annenkovae")), 
        expression(italic("B. elegans")), 
        expression(italic("B. groenlandicus")),
        expression(italic("B. promamme")), 
        "specimen OP473761",
        expression(italic("B. sarsi"))),
                          mylegend = "bottom")
distr_senstr

distr_excl <- distribution(mybasesf = w_boreal_sf,
                          mysf = byl_excl_sf,
                          mycol = colors_excl,
                          mylabels = c(
        expression(italic("B. acutisetis")),
        "Polynoidae gen. sp.", #acutisetis_close
        expression(italic("B. belfastensis")),
        expression(italic("Bylgides sp.")),
        expression(italic("B. fuscus")),
        expression(italic("B. macrolepidus"))),
                          mylegend = "bottom")
distr_excl

distr_acutisetis <- distribution(mybasesf = w_boreal_sf,
                                 mysf = byl_sf_acutisetis,
                                 mycol = colors_acutisetis,
                                 mylabels = c(
        expression(italic("B. acutisetis")),
        "Polynoidae gen. sp."),
                                 mylegend = "bottom")
distr_acutisetis

distr_annenkovae <- distribution(mybasesf = w_arctic_sf,
                                 mysf = byl_sf_annenkovae,
                                 mycol = colors_annenkovae,
                                 mylabels = "none",
                                 mylegend = "none")
distr_annenkovae

distr_elegans <- distribution(mybasesf = w_arctic_sf,
                              mysf = byl_sf_elegans,
                              mycol = colors_elegans,
                              mylabels = "none",
                              mylegend = "none")
distr_elegans

distr_groenlandicus <- distribution(mybasesf = w_arctic_sf,
                                    mysf = byl_sf_groenlandicus,
                                    mycol = colors_groenlandicus,
                                    mylabels = "none",
                                    mylegend = "none")
distr_groenlandicus

distr_promamme <- distribution(mybasesf = w_arctic_sf,
                               mysf = byl_sf_promamme,
                               mycol = colors_promamme,
                               mylabels = "none",
                               mylegend = "none")
distr_promamme

distr_sarsi <- distribution(mybasesf = w_arctic_sf,
                            mysf = byl_sf_sarsi,
                            mycol = colors_sarsi,
                            mylabels = "none",
                            mylegend = "none")
distr_sarsi

```


# check plots for color blindness
```{r}
library(colorBlindness)

# trying the color palette
colors_all <- c("#228834ff", "#f0e442ff", "#d55e00ff", "#7e2954ff","#cc79a7ff","#65dbedff", "#5a8efcff", "#735deeff", "#000049ff", "grey", "#5da899ff", "#ef6678ff")

# test for color blindness
colors_tbl <- tibble(letter = c("a", "a", "b", "b", "b", "c", "d", "d", "d", "e", "e", "e", "e", "f", "f", "f", "g", "g", "g", "g", "h", "h", "i", "i", "i", "j", "j", "j", "k", "k", "l", "l", "l"))
colors_bars <- ggplot(colors_tbl, aes(x = letter, fill = letter)) +
  geom_bar(stat = "count") +
  scale_fill_manual(values = colors_all)
cvdPlot(colors_bars)

cvdPlot(distr_senstr)
cvdPlot(distr_excl)
cvdPlot(distr_acutisetis)
cvdPlot(distr_annenkovae)
cvdPlot(distr_coverage)
cvdPlot(distr_elegans)
cvdPlot(distr_groenlandicus)
cvdPlot(distr_promamme)
cvdPlot(distr_sarsi)
```





# saving maps
```{r}
# make saving function
save_map <- function(out_name, myplot)
  {ggsave(here("output", out_name), plot = myplot, width = 120, height = 80, units = "mm", scale = 1.8)
}

# save all species maps
save_map(out_name = "sampling_coverage.png",
         myplot = distr_coverage)
save_map(out_name = "distribution_sensustricto.png",
         myplot = distr_senstr)
save_map(out_name = "distribution_excludedbylgides.png",
         myplot = distr_excl)
save_map(out_name = "distribution_annenkovae.png",
         myplot = distr_annenkovae)
save_map(out_name = "distribution_elegans.png",
         myplot = distr_elegans)
save_map(out_name = "distribution_groenlandicus.png",
         myplot = distr_groenlandicus)
save_map(out_name = "distribution_promamme.png",
         myplot = distr_promamme)
save_map(out_name = "distribution_sarsi.png",
         myplot = distr_sarsi)
save_map(out_name = "distribution_acutisetis.png",
         myplot = distr_acutisetis)
```


# basic maps
```{r}
europe <- ne_countries(scale = 50, returnclass = "sf", continent = "Europe")

#rnaturalearth
ggplot() +
  geom_sf(data = europe) +
  coord_sf(xlim = c(5, 35), ylim = c(76, 81)) +
  ggtitle("Norden")

#ggOceanMap
basemap(limits = c(5, 35, 76, 81),
        bathymetry = TRUE,
        glaciers = TRUE)

#random map
theme_set(theme_bw())
library("rnaturalearth")
library("rnaturalearthdata")

world <- ne_countries(scale = "medium", returnclass = "sf")

ggplot(data = world) +
  geom_sf(color = "black", fill = "grey") +
  xlab("Longitude") + ylab("Latitude") +
  annotation_scale(location = "bl", width_hint = 0.5) +
    annotation_north_arrow(location = "bl", which_north = "true", 
        pad_x = unit(0.75, "in"), pad_y = unit(0.5, "in"),
        style = north_arrow_fancy_orienteering)
```


# old local distribution map
```{r}
# only use sample data to not screw up projection
byl_local_mod <- byl_locations |> 
  filter(latitude > 71) |> 
  filter(longitude > 0) |> 
  filter(longitude < 30)

# make sf object from sample using its right crs code
byl_sf_local <- st_as_sf(byl_local_mod,
                   coords = c("longitude", "latitude"),
                   crs = 4326)

# make base-map -> try highres here!
europe <- ne_countries(scale = 10, returnclass = "sf", continent = "Europe")


# plot it
ggplot() +
  geom_sf(data = europe) +
  geom_sf(
    data = byl_sf_local,
    aes(color = clade),
    size = 1.5,
    alpha = .8
  ) +
  coord_sf(xlim = c(0, 30), ylim = c(71, 80)) +
  labs(colour = "Clade", x = "Longitude", y = "Latitude", color = "Clade") +
  scale_color_manual(values = c("forestgreen", "orange", "blue1"), labels = c("B. annenkovae", "B. groenlandicus", "B. sarsi")) +
  theme_minimal()
```

# Now make depth plot
```{r}
byl_depths <- byl_raw |> 
  select(clade, depth) |> 
  drop_na() |> 
  filter(clade == "groenlandicus" | clade == "sarsi" | clade == "macrolepidus?" | clade == "annenkovae" | clade == "promamme" | clade == "elegans")

# split the depth column if it is a span
byl_depths <- byl_depths |> 
  mutate(
    depth_start = case_when(
      str_detect(depth, "^\\d+$") ~ as.numeric(depth),
      TRUE ~ as.numeric(str_extract(depth, "^\\d+"))),
    
    depth_end = case_when(
      str_detect(depth, "^\\d+$") ~ as.numeric(depth),
      TRUE ~ as.numeric(str_extract(depth, "(?<=-).*$"))))

# make an average depth for spans and drop rows without depths
byl_depths_av <- byl_depths |> 
  drop_na() |> 
  mutate(depth_av = (depth_start+depth_end)/2)

# make depth categories instead and prepare plot
byl_depths_cat <- byl_depths_av |> 
  mutate(
    depth_category = case_when(
      depth_av >= 0 & depth_av <= 200 ~ "0-200m", 
      depth_av > 200 & depth_av <= 1500 ~ "200-1500m",
      TRUE ~ ">1500m")) |> 
  group_by(clade, depth_category) |> 
  summarise(count = n())
#different categories
byl_depths_cat1 <- byl_depths_av |> 
  mutate(
    depth_category = case_when(
      depth_av >= 0 & depth_av <= 200 ~ "0-200m", 
      depth_av > 200 & depth_av <= 500 ~ "200-500m",
      depth_av > 500 & depth_av <= 2000 ~ "500-2000m",
      TRUE ~ ">2000m")) |> 
  group_by(clade, depth_category) |> 
  summarise(count = n())


# plot it absolute
ggplot(byl_depths_cat, aes(x = depth_category, y = count, fill = clade)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    x = "Depth Category",
    y = "Number of Individuals",
    fill = "Clade"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_viridis_d()
ggplot(byl_depths_cat1, aes(x = depth_category, y = count, fill = clade)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    x = "Depth Category",
    y = "Number of Individuals",
    fill = "Clade"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_viridis_d()


# plot it relative
byl_depths_cat_sum <- byl_depths_cat |> 
  group_by(depth_category) |> 
  summarize(total_count = sum(count))

```
the different sarsi clades do not correspond to depths... the BEST specimens that are separated are from 54 and 102m, and a lot of other sarsi also occurs in the shallow layer.