---
title: "Maps"
output: html_document
date: "2024-12-08"
---

```{r}
library(tidyverse)
library(here)
library(sf)
library(ggspatial)
library(rnaturalearth)
library(ggOceanMaps)
```


# read in data from csv
```{r}
byl_raw <- read.csv(here("Polynoidae - Bylgides mus nub=my num=DNA.csv")) |> janitor::clean_names()
```

# make the right table for locations of specimens
```{r}
byl_locations <- byl_raw |> 
  select(species, clade, martha_s_number, project_number, bold_process_specimen_id, latitude, longitude) |> 
  drop_na() |> 
  filter(clade == "groenlandicus" | clade == "sarsi" | clade == "macrolepidus" | clade == "annenkovae" | clade == "promamme" | clade == "elegans" | clade == "acutisetis") |> 
  mutate(clade = factor(clade)) |> 
  mutate(type = if_else(grepl("type", species, ignore.case = TRUE), "type", "nontype")) |> 
  mutate(type = factor(type))
```

```{r}
byl_sf <- st_as_sf(byl_locations,
                   coords = c("longitude", "latitude"),
                   crs = 4326)
```


# world distribution map (Mercator projection)
```{r}
world <- ne_countries(scale = 10, returnclass = "sf")


# plot it
ggplot() +
  geom_sf(data = world) +
  geom_sf(
    data = byl_sf,
    aes(color = clade),
    size = 1.5,
    alpha = .8
  ) +
  #coord_sf(xlim = c(0, 30), ylim = c(71, 80)) +
  labs(color = "Clade", x = "Longitude", y = "Latitude", color = "Clade") +
  theme_minimal()
```



# world distribution map (polar projection)
```{r}

# load packages
packages <- list("dplyr", "sf", "ggplot2","mapview", "maptools", "raster") #"viridis"
sapply(packages, require, character = TRUE)

# plotting coastlines
data("wrld_simpl", package = "maptools")

w <- raster::crop(wrld_simpl, extent(-180, 180, 30, 90))
    #plot(w)

# make it into an sf object, transform the crs
w_sf <- w |> 
  st_as_sf(coords = c("long", "lat"), crs = 4326) |> 
  st_transform(st_crs(3995))



# make function to plot
distribution <- function(mysf, mycol, mylegend, myshape)
  {ggplot() + 
  geom_sf(data= w_sf, fill = "black", colour = "white", alpha = 0.6) +
  geom_sf(data = mysf, aes(col = clade, shape = type, fill = clade), size = 2, alpha = .9) +
  scale_shape_manual(values = c(19, 24), #19 is a dot, 17 a triangle
                     labels = NULL) + 
  scale_color_manual(values = mycol,
                     labels = c(
        expression(italic("B. acutisetis")),
        expression(italic("B. annenkovae")), 
        expression(italic("B. elegans")), 
        expression(italic("B. groenlandicus")),
        expression(italic("B. macrolepidus")),
        expression(italic("B. promamme")), 
        expression(italic("B. sarsi")))) +
  scale_fill_manual(values = mycol,
                    labels = NULL) +
  scale_x_continuous(breaks = NULL) +
  labs(title = "", x = "", y = "") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank(), 
        text = element_text(size = 12),
        legend.title = element_blank(),
        legend.position = mylegend)
}

```

```{r}
# base R solution
data("wrld_simpl", package = "maptools")

w.equator <- raster::crop(wrld_simpl, extent(-180, 180, 0, 90))  |> 
  st_as_sf(coords = c("long", "lat"), crs = 4326) |> 
  st_transform(st_crs(3995))

donut <- list(cbind(x = c(0:360, 360:0, 0), y = c(rep(45, 361), c(rep(0,361)), 45))) %>%  st_polygon %>%
    st_sfc(crs = 4326) %>%  st_transform(st_crs(3995))

fortyfive <- cbind(x = 0:360, y = rep(45, 361)) %>% st_linestring() %>%  
    st_sfc(crs = 4326) %>% st_transform(st_crs(3995)) 

# putting it together
par(mar = rep(.1,4))
layout(matrix(1:2, ncol =2), widths = c(1.5,1)) 

plot(fortyfive, col = "grey30")
plot(st_geometry(w.equator), col = "white", add = TRUE, border = "grey30")
plot(donut, add = TRUE, col = "white", border = NA)
#plot(lang["LangFamily"], lwd = 1, lty = 1, pal = alpha(cols, .8), add = TRUE, border = NA)

par(mar = c(0,0,6,0), xpd = NA); plot.new()
mtext(side = 3, "Language family", font = 1, line = 0, adj = 0)
legend(x = "topleft", ncol = 1, cex = 0.8, legend = labs, col = cols, pch=15, pt.cex = 1.5, bty = "n")

```

# make sf for species

```{r}
# acutisetis
byl_acutisetis <- byl_locations |> 
 filter(clade == "acutisetis")

byl_sf_acutisetis <- st_as_sf(byl_acutisetis,
                   coords = c("longitude", "latitude"),
                   crs = 4326)

# annenkovae
byl_annenkovae <- byl_locations |> 
  filter(clade == "annenkovae")


byl_sf_annenkovae <- st_as_sf(byl_annenkovae,
                   coords = c("longitude", "latitude"),
                   crs = 4326)

# elegans
byl_elegans <- byl_locations |> 
  filter(clade == "elegans")


byl_sf_elegans <- st_as_sf(byl_elegans,
                   coords = c("longitude", "latitude"),
                   crs = 4326)

# groenlandicus
byl_groenlandicus <- byl_locations |> 
  filter(clade == "groenlandicus")


byl_sf_groenlandicus <- st_as_sf(byl_groenlandicus,
                   coords = c("longitude", "latitude"),
                   crs = 4326)

# promamme
byl_promamme <- byl_locations |> 
  filter(clade == "promamme")


byl_sf_promamme <- st_as_sf(byl_promamme,
                   coords = c("longitude", "latitude"),
                   crs = 4326)

# sarsi
byl_sarsi <- byl_locations |> 
  filter(clade == "sarsi")


byl_sf_sarsi <- st_as_sf(byl_sarsi,
                   coords = c("longitude", "latitude"),
                   crs = 4326)
```

# set colors
```{r}
# set colors for the species and their types
colors_all <- c("#000000ff", "#009e73ff", "#56b4e9ff", "#f0e442ff", "#e69f00ff", "#d55e00ff", "#cc79a7ff")


colors_sampling_coverage <- c("orangered1","orangered1","orangered1","orangered1","orangered1") 
colors_acutisetis <- c("#000000ff", "#000000ff")
colors_annenkovae <- c("#009e73ff","#009e73ff")
colors_groenlandicus <- c("#e69f00ff","#e69f00ff")
colors_macrolepidus <- c("#cc79a7ff", "#cc79a7ff")
colors_promamme <- c("#d55e00ff","#d55e00ff")
colors_elegans <- c("#f0e442ff","#f0e442ff")
colors_sarsi <- c("#56b4e9ff","#56b4e9ff")
```


# plot species distributions
```{r}
distr_coverage <- distribution(mysf = byl_sf,
                               mycol = colors_sampling_coverage,
                               mylegend = "none")
distr_coverage

distr_all <- distribution(mysf = byl_sf,
                          mycol = colors_all,
                          mylegend = "bottom")
distr_all

distr_acutisetis <- distribution(mysf = byl_sf_acutisetis,
                                 mycol = colors_acutisetis,
                                 mylegend = "none")
distr_acutisetis

distr_annenkovae <- distribution(mysf = byl_sf_annenkovae,
                                 mycol = colors_annenkovae,
                                 mylegend = "none")
distr_annenkovae

distr_elegans <- distribution(mysf = byl_sf_elegans,
                              mycol = colors_elegans,
                              mylegend = "none")
distr_elegans

distr_groenlandicus <- distribution(mysf = byl_sf_groenlandicus,
                                    mycol = colors_groenlandicus,
                                    mylegend = "none")
distr_groenlandicus

distr_promamme <- distribution(mysf = byl_sf_promamme,
                               mycol = colors_promamme,
                               mylegend = "none")
distr_promamme

distr_sarsi <- distribution(mysf = byl_sf_sarsi,
                            mycol = colors_sarsi,
                            mylegend = "none")
distr_sarsi

```

# check for color blindness
```{r}
library(colorBlindness)

# trying the color palette
colors_all <- c("#000000ff", "#009e73ff", "#56b4e9ff", "#f0e442ff", "#e69f00ff", "#d55e00ff", "#cc79a7ff")

# test for color blindness
colors_tbl <- tibble(letter = c("a", "a", "b", "b", "b", "c", "d", "d", "d", "e", "e", "e", "e", "f", "f", "f", "g", "g", "g", "g", "h", "h"))
colors_bars <- ggplot(colors_tbl, aes(x = letter, fill = letter)) +
  geom_bar(stat = "count") +
  scale_fill_manual(values = colors_all)
cvdPlot(colors_bars)

cvdPlot(distr_all)
cvdPlot(distr_annenkovae)
cvdPlot(distr_coverage)
cvdPlot(distr_elegans)
cvdPlot(distr_groenlandicus)
cvdPlot(distr_promamme)
cvdPlot(distr_sarsi)
```



# saving maps
```{r}
# make saving function
save_map <- function(out_name, myplot)
  {ggsave(here("output", out_name), plot = myplot, width = 120, height = 80, units = "mm", scale = 1.8)
}

# save all species maps
save_map(out_name = "sampling_coverage2.png",
         myplot = distr_coverage)
save_map(out_name = "distribution_bylgides2.png",
         myplot = distr_all)
save_map(out_name = "distribution_annenkovae2.png",
         myplot = distr_annenkovae)
save_map(out_name = "distribution_elegans2.png",
         myplot = distr_elegans)
save_map(out_name = "distribution_groenlandicus2.png",
         myplot = distr_groenlandicus)
save_map(out_name = "distribution_promamme2.png",
         myplot = distr_promamme)
save_map(out_name = "distribution_sarsi2.png",
         myplot = distr_sarsi)
```


# basic maps
```{r}
europe <- ne_countries(scale = 50, returnclass = "sf", continent = "Europe")

#rnaturalearth
ggplot() +
  geom_sf(data = europe) +
  coord_sf(xlim = c(5, 35), ylim = c(76, 81)) +
  ggtitle("Norden")

#ggOceanMap
basemap(limits = c(5, 35, 76, 81),
        bathymetry = TRUE,
        glaciers = TRUE)

#random map
theme_set(theme_bw())
library("rnaturalearth")
library("rnaturalearthdata")

world <- ne_countries(scale = "medium", returnclass = "sf")

ggplot(data = world) +
  geom_sf(color = "black", fill = "grey") +
  xlab("Longitude") + ylab("Latitude") +
  annotation_scale(location = "bl", width_hint = 0.5) +
    annotation_north_arrow(location = "bl", which_north = "true", 
        pad_x = unit(0.75, "in"), pad_y = unit(0.5, "in"),
        style = north_arrow_fancy_orienteering)
```


# old local distribution map
```{r}
# only use sample data to not screw up projection
byl_local_mod <- byl_locations |> 
  filter(latitude > 71) |> 
  filter(longitude > 0) |> 
  filter(longitude < 30)

# make sf object from sample using its right crs code
byl_sf_local <- st_as_sf(byl_local_mod,
                   coords = c("longitude", "latitude"),
                   crs = 4326)

# make base-map -> try highres here!
europe <- ne_countries(scale = 10, returnclass = "sf", continent = "Europe")


# plot it
ggplot() +
  geom_sf(data = europe) +
  geom_sf(
    data = byl_sf_local,
    aes(color = clade),
    size = 1.5,
    alpha = .8
  ) +
  coord_sf(xlim = c(0, 30), ylim = c(71, 80)) +
  labs(colour = "Clade", x = "Longitude", y = "Latitude", color = "Clade") +
  scale_color_manual(values = c("forestgreen", "orange", "blue1"), labels = c("B. annenkovae", "B. groenlandicus", "B. sarsi")) +
  theme_minimal()
```

# Now make depth plot
```{r}
byl_depths <- byl_raw |> 
  select(clade, depth) |> 
  drop_na() |> 
  filter(clade == "groenlandicus" | clade == "sarsi" | clade == "macrolepidus?" | clade == "annenkovae" | clade == "promamme" | clade == "elegans")

# split the depth column if it is a span
byl_depths <- byl_depths |> 
  mutate(
    depth_start = case_when(
      str_detect(depth, "^\\d+$") ~ as.numeric(depth),
      TRUE ~ as.numeric(str_extract(depth, "^\\d+"))),
    
    depth_end = case_when(
      str_detect(depth, "^\\d+$") ~ as.numeric(depth),
      TRUE ~ as.numeric(str_extract(depth, "(?<=-).*$"))))

# make an average depth for spans and drop rows without depths
byl_depths_av <- byl_depths |> 
  drop_na() |> 
  mutate(depth_av = (depth_start+depth_end)/2)

# make depth categories instead and prepare plot
byl_depths_cat <- byl_depths_av |> 
  mutate(
    depth_category = case_when(
      depth_av >= 0 & depth_av <= 200 ~ "0-200m", 
      depth_av > 200 & depth_av <= 1500 ~ "200-1500m",
      TRUE ~ ">1500m")) |> 
  group_by(clade, depth_category) |> 
  summarise(count = n())
#different categories
byl_depths_cat1 <- byl_depths_av |> 
  mutate(
    depth_category = case_when(
      depth_av >= 0 & depth_av <= 200 ~ "0-200m", 
      depth_av > 200 & depth_av <= 500 ~ "200-500m",
      depth_av > 500 & depth_av <= 2000 ~ "500-2000m",
      TRUE ~ ">2000m")) |> 
  group_by(clade, depth_category) |> 
  summarise(count = n())


# plot it absolute
ggplot(byl_depths_cat, aes(x = depth_category, y = count, fill = clade)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    x = "Depth Category",
    y = "Number of Individuals",
    fill = "Clade"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_viridis_d()
ggplot(byl_depths_cat1, aes(x = depth_category, y = count, fill = clade)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    x = "Depth Category",
    y = "Number of Individuals",
    fill = "Clade"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_viridis_d()


# plot it relative
byl_depths_cat_sum <- byl_depths_cat |> 
  group_by(depth_category) |> 
  summarize(total_count = sum(count))

```
the different sarsi clades do not correspond to depths... the BEST specimens that are separated are from 54 and 102m, and a lot of other sarsi also occurs in the shallow layer.